---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(mosaic)
library(vegan)
source("nes8010.R")
data(dune)
data(dune.env)
data(varespec)
data(varechem)
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction
Sometimes you will have undertaken an ordination where you already know in 
advance that your samples belong to different groups (e.g. different habitats,
or experimental treatments). It can therefore be useful to modify your basic
ordination plots to show these groups or classes. This type of modification to
standard ordination methods is typically applied to unconstrained methods such
as PCA, CA or NMDS.

A second approach is where you have for example species by samples data, and
you wish to classify them on the basis of their species composition. These
classification or 'clustering' methods are widely used in ecology, and they
underpin some of the standard published descriptions of different ecological
communities.

## Modifiying base ordination graphics
If you are using the `ordi_plot()` functions, these produce output in `ggplot`-
compatible format, but can be trickier to modify. You may therefore find it
easier to use the `scores()` function to extract the relevant sample scores,
and then use standard `ggplot` functions.

### Changing point colours
Let's begin with the `dune` and `dune.env` data that you considered earlier.

```{r baseline_pca, exercise=TRUE}
summary(dune.env)  # Notice the three categories for Use
dune.pca <- rda(dune)
ordi_plot(dune.pca, display="sites", type="points")
```

The `dune.env` data splits the management into "Hayfield", "Haypasture" and
"Pasture" and it would be useful to show these as different colour categories
on the plot. The easiest way to do this is to extract PCA scores, and combine
them with the environmental categories for plotting:

```{r tweak_pca_plot-setup}
dune.pca <- rda(dune)
```

```{r tweak_pca_plot, exercise=FALSE, echo=TRUE}
dune_scores <- ordi_scores(dune.pca, display="sites")

ggplot(aes(x = PC1, y=PC2, colour=dune.env$Use), data=dune_scores) +
  geom_point()

```

One disadvantage of this approach is that the legend is captioned "dune.env$Use"
which is a bit messy. You can modify it to something clearer using the
`scale_colour_manual()` function from `ggplot2`. You can also use this function
to manually set the colours to your own choice, and to change the label "Haypastu"
to the clearer "Hay pasture". It is easiest if the `dune.env` column `Use` is
added to the `dune_scores` data.frame, which we can easily do, vor example via
`mutate()`

```{r tweak_pca_plt-setup}
dune.pca <- rda(dune)
```
```{r tweak_pca_plt, exercise=TRUE, echo=TRUE}
dune_scores <- ordi_scores(dune.pca, display="sites")
dune_scores <- mutate(dune_scores, Use=dune.env$Use)

ggplot(aes(x = PC1, y=PC2, colour=Use), data=dune_scores) +
  geom_point() +
  scale_colour_manual(name="Useage", values=c("red","green","blue"), labels=c("Hayfield", "Hay pasture", "Pasture"))

```

**Note** in the above code note that the first line of the `ggplot()` function
uses `colour` rather than `fill` as an option, hence we use `scale_colour_manual`
to fine-tune the appearance. If using `fill` as an option (see below) you will
find it easier to use `scale_fill_discrete` as an option.

### Convex polygons
Another useful way to visulise the ordination plots with categories overlaid is
through minimum convex polygons, in different colours around each class. Here is
and example, again with the dune data:

```{r min_viable_polygon, echo=FALSE}
dune.pca <- rda(dune)
dune_scores <- ordi_scores(dune.pca, display="sites")
dune_scores <- mutate(dune_scores, Use=dune.env$Use)

dune_hull <- dune_scores %>% 
  group_by(Use) %>% 
  slice(chull(PC1, PC2))

ggplot(aes(x = PC1, y=PC2, fill=Use), data=dune_scores) +
  geom_point() +
  scale_fill_discrete(name="Useage", labels=c("Hayfield", "Hay pasture", "Pasture")) +
  geom_polygon(data=dune_hull, alpha=0.5) +
  theme_classic()

```

To create this type of plot we first have to create a **convex hull** which
identifies the outer-most points for each group of your PCA points. We can do
this with `slice()` which cuts out only the relevant rows of your original data,
in combination with the `chull()` function which identifies the points. This is
done by grouping your data on the `Use` variable:

```{r, echo=TRUE}
dune_hull <- dune_scores %>% 
  group_by(Use) %>% 
  slice(chull(PC1, PC2))

print(dune_hull)
```

You can see that `dune_hull` contains fewer rows than the original data, as only
the outermost points are needed for each group. Having created this convex hull,
it can be fed into the `geom_polygon` function. The `alpha` option controls the
transparency: re-run the code below several times to understand its role, by
modifying the value of alpha between 0 and 1.

```{r plot_chull-setup}
dune.pca <- rda(dune)
dune_scores <- ordi_scores(dune.pca, display="sites")
dune_scores <- mutate(dune_scores, Use=dune.env$Use)
```
```{r plot_chull, exercise=TRUE}
# Create the convex hull coordinates
dune_hull <- dune_scores %>% 
  group_by(Use) %>% 
  slice(chull(PC1, PC2))

# Add the polygon to your plot
# Can you remember how to change the theme to classic, as shown above?
ggplot(aes(x = PC1, y=PC2, fill=Use), data=dune_scores) +
  geom_point() +
  scale_fill_discrete(name="Useage", labels=c("Hayfield", "Hay pasture", "Pasture")) +
  geom_polygon(data=dune_hull, alpha=0.5)

```

These examples use 'already known' classes of your samples, where they are
grouped according to useage. However, you can also classify your samples based
on similarity of species compisition. These classes can be shown as **dendrograms**
or overlain in a similar way to what you have just seen onto an ordination plot.

## Multivariate classification or clustering
The vegan package has a number of functions that allow you to do hierarchical classifications of species x samples data tables, and to sort your original tables of data to show the main trends.  Packages such as MASS (default) and labdsv also provide functions for fuzzy clustering and indicator species analysis.  We will focus on the tools available in vegan, using its example "dune" dataset.

### Dissimilarity matrix
Many classification methods are based on measuring 'dissimilarity' between samples, based on the species composition.  Therefore clustering is often a two-step process, involving the construction of a dissimilarity table, and then using an appropriate method to split this table up into separate classes. Here is the dissimilarity table for the dune data; it 
has 20 rows and 20 columns (for the number of quadrats) and the values represent
the disimilarity between 0 and 1 of any pair of quadrats.

```{r dune dist for calc, echo=TRUE}
dune.dis <- vegdist(dune)
print(dune.dis)
```

The table above is a 'triangular matrix' showing the dissimilarity in species composition.  Only half the matrix is needed as this top-right triangle would contain the same information.  Values range between zero and one, with sites that are similar to each other having the lowest values (lowest dissimiliarity).

### Classifying the dissimilarity matrix
There are a number of ways of calculating ecological dissimilarity; the default used by vegdist is the Bray-Curtis index; other common ones you may see used are the Euclidean index and the Gower index (see the help page for the vegdist function for the full list of indices available).
Now that you have calculated the dissimilarity values you can classify and plot the results.  Hierarchical classification produces a dendrogram or tree-like graph to describe the classification; the shape of this graph differs depending on the algorithm used, but I usually find the 'average' linkage method (based on the average dissimilarity between groups of sites) the most robust:

```{r classify dune data, echo=TRUE}
dune.cla <- hclust(dune.dis, method="average")
plot(dune.cla)
```


Reading the dendrogram from top to bottom, the first division splits off samples 14, 15, 16 and 20 from all the others, so that they appear to be in a group on their own.  The next division separates samples 17 and 19 from the remainder.  Then site 1 is in a group of its own.  Finally there are two large groups, one containing sites 3, 4, 8, 9, 12 & 13; the second sites 2, 5, 6, 7, 10, 11 & 18.

Other common methods of clustering are single and complete linkage:

```{r s_and_c-setup}
dune.dis <- vegdist(dune)
```

```{r s_and_c, exercise=TRUE}
dune.cls <- hclust(dune.dis, method="single")
dune.clc <- hclust(dune.dis, method="complete")
plot(dune.cls)
plot(dune.cla)

```

Display the classification trees for these methods and notice how the shape of the trees changes.  Which is the best method?  One simple approach is to measure the correlation between your original dissimilarity matrix and the inter-group dissimilarity at each step up the classification tree via the cophenetic function:

```{r cophenetic-setup}
dune.dis <- vegdist(dune)
dune.cla <- hclust(dune.dis, method="average")
dune.cls <- hclust(dune.dis, method="single")
dune.clc <- hclust(dune.dis, method="complete")
```

```{r cophenetic, exercise=TRUE}
cor(dune.dis, cophenetic(dune.cla))
cor(dune.dis, cophenetic(dune.cls))
cor(dune.dis, cophenetic(dune.clc))
```
Based on the correlations above, which do you think is the best of the three classification methods on these data?

## Deciding on the best number of classes
The exact number of classes you end up with depends both on the classification method used, and the height you decide to divide your dendrogram at.  If the height is set at 0.7 we would have 3 classes for example, but the decision on the number of classes is subjective and your judgement!  Using the complete-linked classification and 3 classes we can do the following
using the `rect.hclust` to display 3 classes on our dendrogram, and `cutree` to
extract the data

```{r, three_groups-setup}
dune.dis <- vegdist(dune)
dune.cla <- hclust(dune.dis, method="average")
dune.cls <- hclust(dune.dis, method="single")
dune.clc <- hclust(dune.dis, method="complete")
```

```{r three_groups, exercise=TRUE}
plot(dune.clc)
rect.hclust(dune.clc, 3)
dune.grp <- cutree(dune.clc, 3)
print(dune.grp)
```

Look at the contents of `dune.grp` and you will see that the `cutree` command produces the classes for each of your samples based on how you have decided to cut your dendrogram into 3 groups.

You can use this classification in the same way as any other factor for your dune vegetation data.  For example, do show how the depth of the A1 soild horizon differs amongst your three classes (__Note:__ include `as.factor(dune.grp)` in the `ggplot` call to ensure that the numeric codes are treated as categorical):

```{r boxplot_grp-setup}
dune.dis <- vegdist(dune)
dune.cla <- hclust(dune.dis, method="average")
dune.cls <- hclust(dune.dis, method="single")
dune.clc <- hclust(dune.dis, method="complete")
dune.grp <- cutree(dune.clc, 3)
```
```{r boxplot_grp, exercise=TRUE}
data(dune.env)

ggplot(dune.env, aes(x=as.factor(dune.grp), y=A1)) +
   geom_boxplot() +
   xlab("Vegetation classification group") +
   ylab("Depth of A1 soil horizon") +
   theme_classic()
```

Of course, depending on your perspective, you might like to think that "explanatory"
variables should be on the horizontal x-axis, and "response" variables on the 
vertical y-axis. So, if you think that the vegetation groups are more likely to
depend on the soil depth, you might want to 'flip' your boxplot. Of course the
decision is up to you, and based on your environmental knowledge! Use `coord_flip()`
to swap axes:

```{r boxplot_grp2-setup}
dune.dis <- vegdist(dune)
dune.cla <- hclust(dune.dis, method="average")
dune.cls <- hclust(dune.dis, method="single")
dune.clc <- hclust(dune.dis, method="complete")
dune.grp <- cutree(dune.clc, 3)
```
```{r boxplot_grp2, exercise=TRUE}
data(dune.env)

ggplot(dune.env, aes(x=as.factor(dune.grp), y=A1)) +
   geom_boxplot() +
   xlab("Vegetation classification group") +
   ylab("Depth of A1 soil horizon") +
   coord_flip() +
   theme_classic()
```



Likewise, you can overlay your classification onto your PCA scores that you
calculated earlier. But now the classes are based on the species, not the habitat::

```{r dune_hull-setup}
dune.pca <- rda(dune)

dune_scores <- ordi_scores(dune.pca, display="sites")

dune.dis <- vegdist(dune)
dune.cla <- hclust(dune.dis, method="average")
dune.cls <- hclust(dune.dis, method="single")
dune.clc <- hclust(dune.dis, method="complete")
dune.grp <- cutree(dune.clc, 3)                           

```
```{r dune_hull, exercise=TRUE}
# Note the use of as.factor to force group numbers to be treated as a factor
dune_scores <- mutate(dune_scores, veg_class=as.factor(dune.grp))

dune_hull <- dune_scores %>% 
  group_by(veg_class) %>% 
  slice(chull(PC1, PC2))

ggplot(aes(x = PC1, y=PC2, fill=veg_class), data=dune_scores) +
  geom_point() +
  scale_fill_discrete(name="Vegetation class") +
  geom_polygon(data=dune_hull, alpha=0.5) +
  theme_classic()

```

The above graph shows a multivariate ordination (via PCA) of your dune data, overlaid with a multivariate classification (via hclust) of the same data.  You also know that the soil depth differs between the three classes, with the highest values (and greatest variability) in Class 2. 
