---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(mosaic)
library(vegan)
source("nes8010.R")
data(dune)
data(dune.env)
data(varespec)
data(varechem)
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction
Sometimes you will have undertaken an ordination where you already know in 
advance that your samples belong to different groups (e.g. different habitats,
or experimental treatments). It can therefore be useful to modify your basic
ordination plots to show these groups or classes. This type of modification to
standard ordination methods is typically applied to unconstrained methods such
as PCA, CA or NMDS.

A second approach is where you have for example species by samples data, and
you wish to classify them on the basis of their species composition. These
classification or 'clustering' methods are widely used in ecology, and they
underpin some of the standard published descriptions of different ecological
communities.

## Modifiying base ordination graphics
If you are using the `ordi_plot()` functions, these produce output in `ggplot`-
compatible format, but can be trickier to modify. You may therefore find it
easier to use the `scores()` function to extract the relevant sample scores,
and then use standard `ggplot` functions.

### Changing point colours
Let's begin with the `dune` and `dune.env` data that you considered earlier.

```{r baseline_pca, exercise=TRUE}
summary(dune.env)  # Notice the three categories for Use
dune.pca <- rda(dune)
ordi_plot(dune.pca, display="sites", type="points")
```

The `dune.env` data splits the management into "Hayfield", "Haypasture" and
"Pasture" and it would be useful to show these as different colour categories
on the plot. The easiest way to do this is to extract PCA scores, and combine
them with the environmental categories for plotting:

```{r tweak_pca_plot-setup}
dune.pca <- rda(dune)
```

```{r tweak_pca_plot, exercise=FALSE, echo=TRUE}
dune_scores <- ordi_scores(dune.pca, display="sites")

ggplot(aes(x = PC1, y=PC2, colour=dune.env$Use), data=dune_scores) +
  geom_point()

```

One disadvantage of this approach is that the legend is captioned "dune.env$Use"
which is a bit messy. You can modify it to something clearer using the
`scale_colour_manual()` function from `ggplot2`. You can also use this function
to manually set the colours to your own choice, and to change the label "Haypastu"
to the clearer "Hay pasture". It is easiest if the `dune.env` column `Use` is
added to the `dune_scores` data.frame, which we can easily do, vor example via
`mutate()`

```{r tweak_pca_plt-setup}
dune.pca <- rda(dune)
```
```{r tweak_pca_plt, exercise=TRUE, echo=TRUE}
dune_scores <- ordi_scores(dune.pca, display="sites")
dune_scores <- mutate(dune_scores, Use=dune.env$Use)

ggplot(aes(x = PC1, y=PC2, colour=Use), data=dune_scores) +
  geom_point() +
  scale_colour_manual(name="Useage", values=c("red","green","blue"), labels=c("Hayfield", "Hay pasture", "Pasture"))

```

**Note** in the above code note that the first line of the `ggplot()` function
uses `colour` rather than `fill` as an option, hence we use `scale_colour_manual`
to fine-tune the appearance. If using `fill` as an option (see below) you will
find it easier to use `scale_fill_discrete` as an option.

### Convex polygons
Another useful way to visulise the ordination plots with categories overlaid is
through minimum convex polygons, in different colours around each class. Here is
and example, again with the dune data:

```{r min_viable_polygon, echo=FALSE}
dune.pca <- rda(dune)
dune_scores <- ordi_scores(dune.pca, display="sites")
dune_scores <- mutate(dune_scores, Use=dune.env$Use)

dune_hull <- dune_scores %>% 
  group_by(Use) %>% 
  slice(chull(PC1, PC2))

ggplot(aes(x = PC1, y=PC2, fill=Use), data=dune_scores) +
  geom_point() +
  scale_fill_discrete(name="Useage", labels=c("Hayfield", "Hay pasture", "Pasture")) +
  geom_polygon(data=dune_hull, alpha=0.5) +
  theme_classic()

```

To create this type of plot we first have to create a **convex hull** which
identifies the outer-most points for each group of your PCA points. We can do
this with `slice()` which cuts out only the relevant rows of your original data,
in combination with the `chull()` function which identifies the points. This is
done by grouping your data on the `Use` variable:

```{r, echo=TRUE}
dune_hull <- dune_scores %>% 
  group_by(Use) %>% 
  slice(chull(PC1, PC2))

print(dune_hull)
```

You can see that `dune_hull` contains fewer rows than the original data, as only
the outermost points are needed for each group. Having created this convex hull,
it can be fed into the `geom_polygon` function. The `alpha` option controls the
transparency: re-run the code below several times to understand its role, by
modifying the value of alpha between 0 and 1.

```{r plot_chull-setup}
dune.pca <- rda(dune)
dune_scores <- ordi_scores(dune.pca, display="sites")
dune_scores <- mutate(dune_scores, Use=dune.env$Use)
```
```{r plot_chull, exercise=TRUE}
# Create the convex hull coordinates
dune_hull <- dune_scores %>% 
  group_by(Use) %>% 
  slice(chull(PC1, PC2))

# Add the polygon to your plot
# Can you remember how to change the theme to classic, as shown above?
ggplot(aes(x = PC1, y=PC2, fill=Use), data=dune_scores) +
  geom_point() +
  scale_fill_discrete(name="Useage", labels=c("Hayfield", "Hay pasture", "Pasture")) +
  geom_polygon(data=dune_hull, alpha=0.5)

```

These examples use 'already known' classes of your samples, where they are
grouped according to useage. However, you can also classify your samples based
on similarity of species compisition. These classes can be shown as **dendrograms**
or overlain in a similar way to what you have just seen onto an ordination plot.